Annotation-Based Empirical Performance Tuning Using Orio

### 2009-02-15 doc/papers/ipdps09/implementation.tex
git-svn-id: https://svn.mcs.anl.gov/repos/performance/orio@428 e7ad4b5f-b827-0410-872f-f7f4bc3d1efb
git-svn-id: https://svn.mcs.anl.gov/repos/performance/orio@433 e7ad4b5f-b827-0410-872f-f7f4bc3d1efb
git-svn-id: https://svn.mcs.anl.gov/repos/performance/orio@443 e7ad4b5f-b827-0410-872f-f7f4bc3d1efb

@@

%\subsection{Overall Approach} 
%\label{sec:overall}
\begin{figure}[tb]
\vspace{-.2in}
\begin{center}  
\includegraphics[width=0.6\textwidth]{figures/orio.eps}   
\end{center}  
\vspace{-.2in}
\caption{Overview of Orio's code generation and empirical tuning process.}  
\label{fig:orio}  
\end{figure}

@@

\begin{figure}[thb]
\begin{center} 
\begin{tabular}{rrl} 
$<$annotation-region$>$ & ::= & $<$leader-annotation$>$ $<$annotation-body$>$ $<$trailer-annotation$>$\\ 
$<$leader-annotation$>$ & ::= & \texttt{/*@ begin} $<$module-name$>$ \texttt{(} $<$module-body$>$ \texttt{) @*/} \\
$<$trailer-annotation$>$ & ::= & \texttt{/*@ end @*/} \\
\end{tabular} 
\end{center}  
\vspace{-.1in}
\caption{Annotation language grammar excerpt.}  
\label{fig:ann-lang}
\end{figure}

@@

\begin{figure}[thb] 
\centering 
\begin{tabular}{cc} 
\begin{minipage}{.5\textwidth}  
\scriptsize
\begin{verbatim}  
void axpy4(int N, double *y, 
 double a1, double *x1, double a2, double *x2, 
 double a3, double *x3, double a4, double *x4) {

@@

\end{verbatim}  
\end{minipage}
&
\begin{minipage}{.5\textwidth}  
\scriptsize
\begin{verbatim}  
spec axpy4_tune_spec {
 def build { 
  arg build_command = 
   'mpixlc_r -O3 -qstrict -qhot -qsmp=omp:noauto'; 
  arg batch_command = 
   'qsub -n 128 -t 20 --env "OMP_NUM_THREADS=4"'; 
  arg status_command = 'qstat';
  arg num_procs = 128; 
 } 
 def performance_counter { 
  arg method = 'basic timer';
  arg repetitions = 10000;
 } 
 def performance_params { 
  param UF[] = range(1,33);
  param PAR[] = [True,False]; 
 }
 def input_params { 
  param N[] = [10,100,1000,10**4,10**5,
               10**6,10**7]; 
 }
 def input_vars { 
  decl dynamic double y[N] = 0; 
  decl dynamic double x1[N] = random; 
  decl double a1 = random; 
  decl double a2 = random; 
  # ... ommitted ...
 } 
 def search { 
  arg algorithm = 'Exhaustive'; 
  arg time_limit = 20;
 }
}
\end{verbatim}  
\end{minipage}
\\
%(a) Annotated AXPY-4 source code
%& 
%(b) Tuning specification for the Blue Gene/P\\
\end{tabular}
\caption{Orio input example; annotated AXPY-4 source code (left) and tuning specification for the Blue Gene/P (right).}
\label{fig:orio-example}  
%\vspace{-.2in}
\end{figure}

@@

As briefly discussed earlier, our empirical tuning approach systematically
measures the performance costs of automatically-generated code variants in
order to find the most optimal available version. In the context of empirical
optimization, code variants are alternative, semantically equivalent,
implementations of the same computation. Each implementation variant is
associated with a collection of different optimization parameters that
correspond to source-to-source code transformations such as, for instance,
unroll factors, tile sizes, and loop permutation order. So, each coordinate
in the search space of empirical tuning problem represents a distinct
combination of performance parameter values. Both the dimension and the size
of the search space depend on the total number and the value ranges of used
performance parameters, respectively.

The conceptually straightforward approach to exploring the space of the
parameter values is to use an exhaustive search procedure that is guaranteed
to determine the optimal code version. Normally the size of the search space
is too large, making full coverage impractical.  Thus, in addition to
supporting exhaustive search, we have implemented several search heuristics.
The simplest search heuristic is a random search, which picks a random
coordinate in the search space at each step and then measures its
performance; random search is not guaranteed to return close-to-optimal results.
We have also developed two other search heuristics to effectively
narrow the search space for close-to-optimal performance, including a
heuristic based on the \textit{Nelder-Mead simplex}
method~\cite{Lagarias98simplex,Lewis00directsearch}, 
a popular non-derivative direct search method for optimization, and \textit{simulated
annealing}\cite{Kirkpatrick83optimizationby}.
%with the additional inclusion of
%a gradually-reduced ``temperature'' used to control the randomness of
%selecting the next move in the search space. 
%By default Orio uses exhaustive
%search, but users can set the search parameters in the tuning specification
%to use and configure one of the available heuristics.  
Similarly to the implementation of Orio's optimization modules, each search
technique is implemented as an independent Python module in Orio and is dynamically
loaded given only the algorithm's name as one of the fields in the tuning
specification.

@@

\begin{figure}[htp] 
\centering 
\begin{tabular}{cc} 
\begin{minipage}{.4\textwidth}  
\scriptsize
\begin{verbatim}  
/*@ begin Loop (
transform UnrollJam(ufactor=Ui)
for (i=0; i<=M-1; i++)
  transform UnrollJam(ufactor=Uj)
  for (j=0; j<=N-1; j++)
    transform UnrollJam(ufactor=Uk)   
    for (k=0; k<=O-1; k++)
      A[i][j] += B[i][k]*C[k][j];
) @*/
for (i=0; i<=M-1; i++)
 for (j=0; j<=N-1; j++)
  for (k=0; k<=O-1; k++)
    A[i][j] += B[i][k]*C[k][j];
/*@ end @*/
\end{verbatim}  
\end{minipage}
&
\begin{minipage}{.45\textwidth}  
\scriptsize
\begin{verbatim}  
def performance_params {
  param Ui[] = range(1,33);
  param Uj[] = range(1,33);
  param Uk[] = range(1,33);
  constraint reg_capacity = Ui*Uj+Ui*Uk+Uk*Uj<=32;
}

\end{verbatim}  
\end{minipage}
\\
%(a) Annotated code of matrix-matrix multiplication
%& 
%(b) Orio's parameter specifications\\
\end{tabular}
\caption{Example of specifying parameter constraints in Orio; annotated code for matrix-matrix multiplication (left) and constraint specification (right).}
\label{fig:par-constraint}  
\end{figure}

>>>>

\begin{figure*}%[b]
\begin{center}
\includegraphics[width=0.6\textwidth]{figures/orio.eps}   
\end{center}
\caption{Overview of Orio's code generation and empirical tuning process.}  
\label{fig:orio}
\end{figure*}

\begin{figure*}%[b]
\begin{center} 
\begin{tabular}{rrl} 
$<$annotation-region$>$ & ::= & $<$leader-annotation$>$ $<$annotation-body$>$ $<$trailer-annotation$>$\\ 
$<$leader-annotation$>$ & ::= & \texttt{/*@ begin} $<$module-name$>$ \texttt{(} $<$module-body$>$ \texttt{) @*/} \\
$<$trailer-annotation$>$ & ::= & \texttt{/*@ end @*/} \\
\end{tabular} 
\end{center}  
\caption{Annotation language grammar excerpt.}  
\label{fig:ann-lang}
\end{figure*}

@@

@@

@@

\begin{figure*}%[!t] 
\centering 
\begin{tabular}{cc} 
\begin{minipage}{.5\textwidth}  
\scriptsize
\begin{verbatim}  
 void axpy4(int N, double *y, 
  double a1, double *x1, double a2, double *x2, 
  double a3, double *x3, double a4, double *x4) {

@@

\end{verbatim}  
\end{minipage}
&
\begin{minipage}{.5\textwidth}  
\scriptsize
\begin{verbatim}  
 spec axpy4_tune_spec {
  def build { 
   arg build_command = 
    'mpixlc_r -O3 -qstrict -qhot -qsmp=omp:noauto'; 
   arg batch_command = 
    'qsub -n 128 -t 20 --env "OMP_NUM_THREADS=4"'; 
   arg status_command = 'qstat';
   arg num_procs = 128; 
  } 
  def performance_counter { 
   arg method = 'basic timer';
   arg repetitions = 10000;
  } 
  def performance_params { 
   param UF[] = range(1,33);
   param PAR[] = [True,False]; 
  }
  def input_params { 
   param N[] = [10,100,1000,10**4,10**5,
                10**6,10**7]; 
  }
  def input_vars { 
   decl dynamic double y[N] = 0; 
   decl dynamic double x1[N] = random; 
   decl double a1 = random; 
   decl double a2 = random; 
   # ... ommitted ...
  } 
  def search { 
   arg algorithm = 'Exhaustive'; 
   arg time_limit = 20;
  }
 }
\end{verbatim}  
\end{minipage}
\\
\end{tabular}
\caption{Orio input example; annotated AXPY-4 source code (left) and tuning specification for the Blue Gene/P (right).}
\label{fig:orio-example}  
\end{figure*}

\begin{figure*}%[!t] 
\centering 
\begin{tabular}{cc} 
\begin{minipage}{.5\textwidth}  
\scriptsize
\begin{verbatim}  
 /*@ begin Loop (
 transform UnrollJam(ufactor=Ui)
 for (i=0; i<=M-1; i++)
   transform UnrollJam(ufactor=Uj)
   for (j=0; j<=N-1; j++)
     transform UnrollJam(ufactor=Uk)   
     for (k=0; k<=O-1; k++)
       A[i][j] += B[i][k]*C[k][j];
 ) @*/
 for (i=0; i<=M-1; i++)
  for (j=0; j<=N-1; j++)
   for (k=0; k<=O-1; k++)
     A[i][j] += B[i][k]*C[k][j];
 /*@ end @*/
\end{verbatim}  
\end{minipage}
&
\begin{minipage}{.5\textwidth}  
\scriptsize
\begin{verbatim}  
 def performance_params {
   param Ui[] = range(1,33);
   param Uj[] = range(1,33);
   param Uk[] = range(1,33);
   constraint reg_capacity = Ui*Uj+Ui*Uk+Uk*Uj<=32;
 }

\end{verbatim}  
\end{minipage}
\\
\end{tabular}
\caption{Example of specifying parameter constraints in Orio; annotated code for matrix-matrix multiplication (left) and constraint specification (right).}
\label{fig:par-constraint}  
\end{figure*}

@@

As briefly discussed earlier, our empirical tuning approach systematically
measures the performance costs of automatically-generated code variants in
order to find the most optimal available version. In the context of empirical
optimization, code variants are alternative, semantically equivalent
implementations of the same computation. Each implementation variant is
associated with a collection of different optimization parameters that
correspond to source-to-source code transformations such as, for instance,
unroll factors, tile sizes, and loop permutation order. So, each coordinate
in the search space of empirical tuning problem represents a distinct
combination of performance parameter values. Both the dimension and the size
of the search space depend on the total number and the value ranges of used
performance parameters, respectively.

The conceptually straightforward approach to exploring the space of the
parameter values is to use an exhaustive search procedure that is guaranteed
to find the optimal code version. Normally the size of the search space
is too large, making full coverage impractical.  Thus, in addition to
supporting exhaustive search, we have implemented several search heuristics.
The simplest search heuristic is a random search, which picks a random
coordinate in the search space at each step and then measures its
performance; random search is not guaranteed to return close-to-optimal results.
We have also developed two other search heuristics to effectively
narrow the search space for close-to-optimal performance, including a
heuristic based on the \textit{Nelder-Mead simplex}
method~\cite{Lagarias98simplex,Lewis00directsearch}, 
a popular non-derivative direct search method for optimization, and \textit{simulated
annealing}\cite{Kirkpatrick83optimizationby}.
%with the additional inclusion of
%a gradually-reduced ``temperature'' used to control the randomness of
%selecting the next move in the search space. 
%By default Orio uses exhaustive
%search, but users can set the search parameters in the tuning specification
%to use and configure one of the available heuristics.  
Similarly to the implementation of Orio's optimization modules, each search
technique is implemented as an independent Python module in Orio and is dynamically
loaded given only the algorithm's name as one of the fields in the tuning
specification.

@@
